Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-01-23 papers/histogram/sad.tex

Transition Matrix Monte Carlo~\cite{wang1999transition,
swendsen1999transition, fitzgerald2000monte} became an attractive
complimentary simulation algorithm to Wang-Landau, in that, if detailed
balance is ensured, the system is guaranteed to converge.  Unfortunately,
the algorithm can be considerably more difficult to implement when
compared to Wang-Landau: in particular, due to the infinite temperature
transition matrix~\cite{wang2002transition}.  Also, the time necessary
for the density of states to converge could be considerable
due to the algorithm spending too much time exploring low energy states.
Unlike WL, TMMC does not require a prior knowledge of the energy range of
the system.

Shell et al.~\cite{shell2003improved, shell2004flat} originially
implemented Wang-Landau Transition Matrix Monte Carlo (WL-TMMC) in an
effort to quickly explore the energies of a given system using WL then
switch over to TMMC to guarantee convergence. Considerable effort has
been spent trying to determine the flatness criteria and cutoff for the
WL portion, but there remains no rigorous systematic way to determine
these parameters~\cite{rane2013monte}.  The algorithm is first run for
several trial simulations and/or `experience' is used to determine the
best parameters for a users particular system of
study~\cite{siderius2013use}.

Around the same time, the convergence rate for WL was being
explored~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. It was found that utilizing modification factors
that decrease faster than $1/t^2$ leads to
nonconvergence~\cite{belardinelli2007fast}.  This leads to the so-called
$1/t$-WL algorithm which ensures that excess CPU time is not wasted by
continuing to perform calculations once the error in the density of
states becomes saturated~\cite{belardinelli2008analysis}. An effort to
reduce the number of user defined parameters of WL and formulate a
theory for why the method converges despite detailed balance being
violated although infrequently was also being undertaken.

@@

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamic
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several flat histogram methods
which include TMMC, WL, WL-TMMC, and SAMC.

@@

The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
describe in detail flat-histogram methods.  Section~\ref{sec:weight} outlines
weight-based histogram methods used in this work. Section~\ref{sec:transition}
details the workings of the transition-matrix methods studied in this work.
Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
In Section~\ref{sec:results}, we discuss the performance of the various methods
applied to the model system.

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Transition Matrix Monte-Carlo (TMMC), Wang-Landau
(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
(SAD).

@@

\section{Transition-matrix methods}\label{sec:transition}

\subsection{TMMC Algorithm}
The previously described weight-based flat-histogram methods utilizes a so-called
``visited states'' approach to calculate  thermodynamic
properties~\cite{errington2003direct}. The algorithm is fundamentally
based on the number of times a given state is visited while the
system is being sampled. The Transition Matrix Monte Carlo (TMMC) method operates
differently than the visited states approach~\cite{fitzgerald2000monte}.
The method uses attempted transitions between states to obtain an estimate of
the probability that a given transition will be proposed.  The density
of states is computed from the resulting transition matrix.  An advantage
of collecting the transition matrix is that this matrix may be collected
while another method is employed to compute acceptance probabilities, as
in Wang-Landau TMMC, discussed below.

The TMMC algorithm records the transition data in a \emph{collection matrix},
i.e. the unnormalized transition histogram matrix, which stores the number
of times a transition between two energy states has been proposed.
\begin{align}
  C(E_{\text{old}} \rightarrow E_{\text{new}},t+1)
     = C(E_{\text{old}} \rightarrow E_{\text{new}},t) + 1
\end{align}
The transition matrix is computed from the collection matrix via
\begin{align}
  T(E_{\text{old}} \rightarrow E_{\text{new}})
   = \frac{C(E_{\text{old}} \rightarrow E_{\text{new}})}{
     \sum_E C(E_{\text{old}} \rightarrow E)
  }
\end{align}
The acceptance probability is defined in terms of this biasing function:
\begin{align}
  \mathcal{P}(E_{\text{old}} \rightarrow E_{\text{new}}) =
  \max\left(1,\frac{T(E_{\text{old}} \rightarrow E_{\text{new}})}{
       T(E_{\text{new}} \rightarrow E_{\text{old}})}\right)
\end{align}
where we note that the transition matrix is updated to reflect the
transition under consideration \emph{prior} to computing the acceptance
probability.  The TMMC method suffers from poor convergence primarily
because it is sensitive to poor statistics in less probable energy
transitions.  Like the weight-based methods above, the TMMC method violates
detailed balance, with the level of violation decreasing as the simulation
progresses and a difference of 1 in the collection matrix has a smaller
impact on the acceptance probabilities.

\subsection{WL-TMMC Algorithm}
Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC) method as proposed
by Shell and coworkers~\cite{shell2003improved,shell2004flat} takes
advantage of the WL methods ease of implementation, while making use of
a transition matrix to address WL weaknesses, such as its limited
statistical accuracy convergence.

\paragraph{Initializing Wang-Landau:}
We set the energy range along which the method may sample as
$E(T_{\min}) <E< E(T=\infty)$. Conventional Wang-Landau prescribes
updating the WL factor $\gamma$ by starting at 1.0 and reducing by
$\frac12$ until the simulation is cutoff typically around $10^{-10}$.
While variants of this approach appear in the literature, Rane et. al.
published an implemenation of WL-TMMC~\cite{rane2013monte} that
suggested using a WL factor between 1.0 and $10^{-2}$ with a cutoff
$<10^{-5}$, We implement WL-TMMC by setting the WL factor to be 1.0 and
the cutoff to be either $10^{-4}$ or $10^{-10}$.

A single sweep ``roundtrip'' is defined as each macrostate being
sampled a number of times in this case 100,000 times. Wang-Landau sets
a flatness criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually to be 0.8.  In the original implementation, it was
found to be sufficient that each macrostate is visited at least once
(each histogram bin has one entry)~\cite{shell2003improved}; however,
for other systems this is rarely sufficient~\cite{bhar2009computer}.
The goal of this WL initialization is to prefill the transition matrix
for the next portion of the simulation.

\paragraph{Initializing TMMC:} Prefilling the transition matrix is
necessary because if numerous zeros existed in the collection matrix,
the infinite temperature transition matrix would be ill-defined.
Sampling over large ranges of density of states would therefore take an
unacceptable length of time to complete~\cite{shell2003improved,
shen2014elucidating}.  The TMMC portion of the algorithm is run only
after WL has reached the specified cutoff. An important note is that
the acceptatnce probability for TMMC is only used to control actual
transitions not update the collection matrices.

@@

As a side note, WL-TMMC does not converge well for $\delta_0 = 0.05\sigma$ while
it does perform well for $\delta_0 = 0.005\sigma$. The WL portion of the algorithm
spends nearly 100x longer for the smaller translation scale ensuring that the algorithm
converges quickly upon transitioning to TMMC.
}

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that
were carried out that all the methods from the weight-based family
struggle to converge if the update factor decreases too rapidlly.
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} shows SAMC with inappropriate choices
of $t_0$ i.e. less than $\nicefrac{10^5}{\sqrt{t}}$ failing to converge
in a reasonable amount of time. Thus, the update factor $\gamma$, shown
as a function of simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves
useful in determining which methods will converge in a reasonable
amount of time. We compare each method to a reference SAMC
simulation for an ideal choice of $t_0$.  This gives an unbiased
comparison with our developed method SAD. The dynamic version of SAMC
which we call SAD performs as well as SAMC (assuming the user can
correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

@@

We have introduced a new algorithm SAD (Stochastic Approximation with a
Dynamic $\gamma$) that effectively samples the entire energy space for
a variety of system sizes.  Also, SAD proves effective in comparison
with other Monte-Carlo methods at converging to the correct density of
states.  The ease of implementation and the absence of user-defined
parameters makes SAD an ideal algorithm. SAD converges as quickly as
SAMC (in the case of an ideally chosen $t_0$ for our simulations. SAD
doesn't suffer from the weakness of SAMC, i.e. the user choosing
parameters that could lead to the simulation not converging in a
reasonable amount of time. SAD also converges more quickly than Wl,
TMMC, and WL-TMMC.

@@

We gratefully acknowledge D.W. Siderius and
J.R. Errington for helpful communications regarding the implementation
of WL-TMMC.

>>>>

%Transition Matrix Monte Carlo~\cite{wang1999transition,
%swendsen1999transition, fitzgerald2000monte} became an attractive
%complimentary simulation algorithm to Wang-Landau, in that, if detailed
%balance is ensured, the system is guaranteed to converge.  Unfortunately,
%the algorithm can be considerably more difficult to implement when
%compared to Wang-Landau: in particular, due to the infinite temperature
%transition matrix~\cite{wang2002transition}.  Also, the time necessary
%for the density of states to converge could be considerable
%due to the algorithm spending too much time exploring low energy states.
%Unlike WL, TMMC does not require a prior knowledge of the energy range of
%the system.
%
%Shell et al.~\cite{shell2003improved, shell2004flat} originially
%implemented Wang-Landau Transition Matrix Monte Carlo (WL-TMMC) in an
%effort to quickly explore the energies of a given system using WL then
%switch over to TMMC to guarantee convergence. Considerable effort has
%been spent trying to determine the flatness criteria and cutoff for the
%WL portion, but there remains no rigorous systematic way to determine
%these parameters~\cite{rane2013monte}.  The algorithm is first run for
%several trial simulations and/or `experience' is used to determine the
%best parameters for a users particular system of
%study~\cite{siderius2013use}.
\jpsays{I should look at some more 1/t-WL citations. I know a few more
that probably should be included here...}
Around the same time, the convergence rate for WL was being
explored~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. It was found that utilizing modification factors
that decrease faster than $1/t^2$ leads to
nonconvergence~\cite{belardinelli2007fast}.  This leads to the so-called
$1/t$-WL algorithm which ensures that excess CPU time is not wasted by
continuing to perform calculations once the error in the density of
states becomes saturated~\cite{belardinelli2008analysis}. An effort to
reduce the number of user defined parameters of WL and formulate a
theory for why the method converges despite detailed balance being
violated although infrequently was also being undertaken.

@@

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamic
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several flat histogram methods
which include WL, 1/t-WL, and SAMC.
%which include TMMC, WL, WL-TMMC, and SAMC.

@@

The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
describe in detail flat-histogram methods.  Section~\ref{sec:weight} outlines
weight-based histogram methods used in this work.
%Section~\ref{sec:transition}
%details the workings of the transition-matrix methods studied in this work.

Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
In Section~\ref{sec:results}, we discuss the performance of the various methods
applied to the model system.

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Wang-Landau (WL), 1/t-Wang-Landau (1/t-WL), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation (SAD).

%Transition Matrix Monte-Carlo (TMMC), Wang-Landau
%(WL), Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC), Stochastic
%Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation
%(SAD).

@@

%\section{Transition-matrix methods}\label{sec:transition}
%
%\subsection{TMMC Algorithm}
%The previously described weight-based flat-histogram methods utilizes a so-called
%``visited states'' approach to calculate  thermodynamic
%properties~\cite{errington2003direct}. The algorithm is fundamentally
%based on the number of times a given state is visited while the
%system is being sampled. The Transition Matrix Monte Carlo (TMMC) method operates
%differently than the visited states approach~\cite{fitzgerald2000monte}.
%The method uses attempted transitions between states to obtain an estimate of
%the probability that a given transition will be proposed.  The density
%of states is computed from the resulting transition matrix.  An advantage
%of collecting the transition matrix is that this matrix may be collected
%while another method is employed to compute acceptance probabilities, as
%in Wang-Landau TMMC, discussed below.
%
%The TMMC algorithm records the transition data in a \emph{collection matrix},
%i.e. the unnormalized transition histogram matrix, which stores the number
%of times a transition between two energy states has been proposed.
%\begin{align}
%  C(E_{\text{old}} \rightarrow E_{\text{new}},t+1)
%     = C(E_{\text{old}} \rightarrow E_{\text{new}},t) + 1
%\end{align}
%The transition matrix is computed from the collection matrix via
%\begin{align}
%  T(E_{\text{old}} \rightarrow E_{\text{new}})
%   = \frac{C(E_{\text{old}} \rightarrow E_{\text{new}})}{
%     \sum_E C(E_{\text{old}} \rightarrow E)
%  }
%\end{align}
%The acceptance probability is defined in terms of this biasing function:
%\begin{align}
%  \mathcal{P}(E_{\text{old}} \rightarrow E_{\text{new}}) =
%  \max\left(1,\frac{T(E_{\text{old}} \rightarrow E_{\text{new}})}{
%       T(E_{\text{new}} \rightarrow E_{\text{old}})}\right)
%\end{align}
%where we note that the transition matrix is updated to reflect the
%transition under consideration \emph{prior} to computing the acceptance
%probability.  The TMMC method suffers from poor convergence primarily
%because it is sensitive to poor statistics in less probable energy
%transitions.  Like the weight-based methods above, the TMMC method violates
%detailed balance, with the level of violation decreasing as the simulation
%progresses and a difference of 1 in the collection matrix has a smaller
%impact on the acceptance probabilities.
%
%\subsection{WL-TMMC Algorithm}
%Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC) method as proposed
%by Shell and coworkers~\cite{shell2003improved,shell2004flat} takes
%advantage of the WL methods ease of implementation, while making use of
%a transition matrix to address WL weaknesses, such as its limited
%statistical accuracy convergence.
%
%\paragraph{Initializing Wang-Landau:}
%We set the energy range along which the method may sample as
%$E(T_{\min}) <E< E(T=\infty)$. Conventional Wang-Landau prescribes
%updating the WL factor $\gamma$ by starting at 1.0 and reducing by
%$\frac12$ until the simulation is cutoff typically around $10^{-10}$.
%While variants of this approach appear in the literature, Rane et. al.
%published an implemenation of WL-TMMC~\cite{rane2013monte} that
%suggested using a WL factor between 1.0 and $10^{-2}$ with a cutoff
%$<10^{-5}$, We implement WL-TMMC by setting the WL factor to be 1.0 and
%the cutoff to be either $10^{-4}$ or $10^{-10}$.
%
%A single sweep ``roundtrip'' is defined as each macrostate being
%sampled a number of times in this case 100,000 times. Wang-Landau sets
%a flatness criteria~\cite{wang2001determining, wang2001efficient,
%hatch2015computational, mahynski2017predicting} for the accumulated
%histogram usually to be 0.8.  In the original implementation, it was
%found to be sufficient that each macrostate is visited at least once
%(each histogram bin has one entry)~\cite{shell2003improved}; however,
%for other systems this is rarely sufficient~\cite{bhar2009computer}.
%The goal of this WL initialization is to prefill the transition matrix
%for the next portion of the simulation.
%
%\paragraph{Initializing TMMC:} Prefilling the transition matrix is
%necessary because if numerous zeros existed in the collection matrix,
%the infinite temperature transition matrix would be ill-defined.
%Sampling over large ranges of density of states would therefore take an
%unacceptable length of time to complete~\cite{shell2003improved,
%shen2014elucidating}.  The TMMC portion of the algorithm is run only
%after WL has reached the specified cutoff. An important note is that
%the acceptatnce probability for TMMC is only used to control actual
%transitions not update the collection matrices.

@@

%As a side note, WL-TMMC does not converge well for $\delta_0 = 0.05\sigma$ while
%it does perform well for $\delta_0 = 0.005\sigma$. The WL portion of the algorithm
%spends nearly 100x longer for the smaller translation scale ensuring that the algorithm
%converges quickly upon transitioning to TMMC.
}

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that
were carried out that all the methods from the weight-based family
struggle to converge if the update factor decreases too rapidlly.
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} shows SAMC with inappropriate choices
of $t_0$ i.e. less than $\nicefrac{10^5}{\sqrt{t}}$ failing to converge
in a reasonable amount of time. Thus, the update factor $\gamma$, shown
as a function of simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves
useful in determining which methods will converge in a reasonable
amount of time. We compare each method to a reference SAMC
simulation for an ideal choice of $t_0$.  This gives an unbiased
comparison with our developed method SAD. The dynamic version of SAMC
which we call SAD performs as well as SAMC (assuming the user can
correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and 1/t-WL? for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

@@

We have introduced a new algorithm SAD (Stochastic Approximation with a
Dynamic $\gamma$) that effectively samples the entire energy space for
a variety of system sizes.  Also, SAD proves effective in comparison
with other Monte-Carlo methods at converging to the correct density of
states.  The ease of implementation and the absence of user-defined
parameters makes SAD an ideal algorithm. SAD converges as quickly as
SAMC (in the case of an ideally chosen $t_0$ for our simulations. SAD
doesn't suffer from the weakness of SAMC, i.e. the user choosing
parameters that could lead to the simulation not converging in a
reasonable amount of time. 
%SAD also converges more quickly than Wl,
%TMMC, and WL-TMMC.

@@

%We gratefully acknowledge D.W. Siderius and
%J.R. Errington for helpful communications regarding the implementation
%of WL-TMMC.
